Idealized Wind Models of Cool Gas Outflows

### 2010-08-16 Winds/papers/Theory/First/wind_simple.tex
profx
git-svn-id: https://svn.ucolick.org/xsvn/trunk/radtrans@5130 ac11df5e-a788-c23b-a8e1-c758c323924d

Dust is frequently invoked to explain the suppression of line emission
related to a resonantly trapped line (e.g.\ \lya).  Indeed, a photon
that is trapped for many scatterings within a dust medium will be
preferentially extincted relative to a non-resonant photon.  In
$\S$~\ref{sec:dust}, we examined the effects of dust extinction on the
\ion{Mg}{2} and \ion{Fe}{2} emission.  The general result was a modest
suppression of the line flux of the order of (1+\taud)$^{-1}$ instead of
exp(--\taud).   [Do a series of models and fit a functional form for
the fiducial model]  Although the \ion{Mg}{2} photons are resonantly
trapped within the outflow, they require only one to a few scatterings
to escape the fiducial wind model and dust has only a
modest effect.  This reflects the [modest] opacity of the \ion{Mg}{2}
doublet (e.g.\ relative to \lya) but also the specific characteristics
of the wind.  We have also experimented with dust in several of the
other models (e.g.\ the ISM+wind scenario) and find similar results.
[Dust does reduce the $v \sim 0$ emission in the ISM+wind model more,
as it is more effectively trapped.  Discuss this somewhere..  Also
note that the EW of the absorption is higher for an ISM model so that
the absolute value of EW emission is higher too for FeII*]
Another reason \ion{Mg}{2} is less susceptible than \lya\ is that the
latter may be generated through the flourescence of an ionizing
photon, i.e.\ via a recombination cascade.  The ionizing photons 
would penetrate deeper into the outflow and more effectively trap the
\lya\ photon [is this right?].

We also emphasize that the \ion{Fe}{2} resonance photons, especially
\feiia, are converted to non-resonant photons and escape the wind
freely.  A wind model that effectively trapped \ion{Mg}{2}
photons for many scatterings would not be expected to trap the
\feiid\ photons.  Therefore, the effects of dust are 
reduced for \ion{Fe}{2}* and, by inference, any other set of
transitions that are coupled to a fine-structure level (e.g.\
\ion{Si}{2} and \ion{Si}{2}*).  
Although dust does reduce the line flux relative to the continuum, we
conclude that it has only a modest effect on \ion{Mg}{2} and
\ion{Fe}{2}* emission.

Another factor that reduces the predicted line emission is a
non-isotropic wind model,e.g.\ the flux is reduced if there is no
backside of the wind.  To zeroth order, the line emission scales
roughly as $\Omega/4\pi$ where $\Omega$ is the angular extent of the
wind\footnote{There is also a first order dependence on the actual
  orientation of the wind with emission maximal if the backside is
  fully present.}.  Because we require that the wind is at least
pointing towards us, it is difficult to reduce $\Omega$ signficantly
below $2 \pi$.  Therefore, we expect this effect to reduce the
emission by a factor of order unity.  
[The notion of the source absorbing all photons that scatter into it
is akin to removing the backside]

By a similar [token], an anisotropic source could alter the predicted
line emission.  For example, if the backside of the galaxy were
brighter/fainter than would one predict brighter/fainter emission
relative to the observed continuum.
To reduce the emission, the brightest regions
of the galaxy would need to be oriented toward Earth.  Although this would not a
generic occurance, most spectroscopic samples are magnitude-limited
and therefore biased to detecting galaxies at their brightest.

Together, dust and non-isotropic wind models can reduce line emission
by a factor of a few.  We suggest, however, that this would generally
not be sufficient to reduce the emission below the typical detection
limits.

There is another, more subtle effect that in principle will
substantially reduce the observed line emission: slit losses.  As
described in Figures~\ref{fig:fiducial_ifu_mgii} and
\ref{fig:ISM_ifu}, emission from the wind is spatially extended with
at least modest surface brightness at large radii implying a
non-negligible total luminosity from well beyond the galaxy.  
The majority of observations of star-forming galaxies to date have
been taken through a spectroscopic slit typically designed to cover
the brightest continuum regions of the galaxy.  A slit with an angular
extent of $1-1.5''$ subtends roughly $8-15$\,kpc for galaxies at $z>0.5$.
Therefore, a slit covering a galaxy with our fiducial wind would cover
less than half of the wind.  The result will be reduced line emission,
especially in comparison with the galaxy continuum.

We have investigated the effect of a [partial] aperture as follows.
We model the slit as a perfect, infinte rectangle centered on the
galaxy.  We then tabulate the equivalent width of the line emission
through slits with a range of widths, parameterized by the radius
where the wind has an effective optical depth in the line of $\tau =
0.2$ (\rtt)\footnote{Specifically, we determine the
  velocity where $\tau(v) = 0.2$ and then use the velocity law to
  convert to a radius.}.   We expect the line emission is weak beyond
this radius because the photons have only a low probability of
interacting with the wind.  For the fiducial wind model, $\mrtt
\approx 15$\,kpc for \mgiia\ (Figure~\ref{fig:fiducial_nvt}).

The predicted equivalent width in emission \ewe\ relative to the absorption
equivalent width \ewabs\ is presented in
Figure~\ref{fig:obs_slit} for a set of transitions (black curves).
The curves rise very steeply with increasing slit width and then
plateau when the slit width is $\approx \mrtt$.  
The results are comparable for other wind models considered
in this paper (colored curves in Figure~\ref{fig:obs_slit}).
We conclude that slit losses would be important for slits that do not
cover a majority of the wind.   Again, the effect is of order unity
unless the slit width is very small.   Provided the winds do not
extend far beyond 10\,kpc, the majority of emission will be subtended
by stnadard slit observations.  Nevertheless, the results do motivate
the usage of extended apertures, e.g.\ integral field unit (IFU)
instrumentation that would map the wind both spatially and spectrally.

Although there are several effects that will reduce line emission
relative to the absorption produced by the outflow, our analysis
indicates that detectable line emission should be present for [the
majority] of winds.  One of the principal effects of this emission is
that it modifies the observed absorption profiles (e.g.\
Figure~\ref{fig:noemiss}) which may lead to erroneous conclusions
regarding characteristics of the flow.  {\it Analysis of cool gas outflows
that ignores emission may incorrectly lead to conclusions the gas is
partially covered, has a significantly lower peak optical depth,
and/or that a $v \sim 0 \mkms$ ISM component (perhaps even infall) is
absent.}  Figure~\ref{fig:obs_ew} demonstrates a simple observational
consequence of the line emission: reduced estimates for the absorption
equivalent width \ewabs\ of the flow.  In the case of \ion{Mg}{2},
which has the most affected transitions, the \ewabs\ is reduced by
$30-50\%$ for the wind models presented in this paper.  This could
yield a systematically lower optical depth for the wind and therefore
a lower total mass and kinetic energy.  Although not apparent in
Figure~\ref{fig:obs_ew}, the effects are most pronounced for wind
scenarios where the peak optical depth occurs near $v = \mkms$.
Simple geometric projection limits the majority of scattered photon emission to
have $v \approx \pm 200 \mkms$ and the absorption profiles
are filled-in primarily at these velocities.  In
Figure~\ref{fig:obs_peaktau}, we plot the observed peak optical depth
for \mgiia\ 
versus the velocity where $\tau = \tau_{pk}$ for the various wind models (i.e.\
$\tau_{pk}$ vs.\ $v_\tau$ from Tables~\ref{tab:line_diag},
\ref{tab:plaw_diag}).   In every one of the models, the intrinsic peak
optical depth exceeds 10 and occasionally exceeds 100.  For the
majority of cases with $v_\tau < -300\mkms$, one observes $\tau_{pk} <
2$ and would infer the wind is not even optically thick!
The results are similar for \mgiib\ although one infers a somewhat
higher column density because of its lower $f\lambda$ value.
In the majority of cases, one would faslely conclude that hte wind
partially covered the source and also estimate a peak optical depth
lower than the intrinsic.

All of these (negative) effects for \ion{Mg}{2} absorption are reduced
for the \ion{Fe}{2} absorption profiles, especially for \feiia.  This
is because a signficant fraction (or even a majority) of the
\ion{Fe}{2} emission is florescent \ion{Fe}{2}* emission at longer
wavelengths which have no effect on the absorption profiles. [word]
Therefore, the \ion{Fe}{2} absorption equivalent widths (\ewabs) more
closely follow the intrinsic value, one derives more accurate peak
optical depths, and the absorption kinematics more faithfully reflect
the motions of the flow.  Regarding the last point, one 
predicts a velocity offset between the peak depth of the \ion{Fe}{2}
centroid and that for \ion{Mg}{2} (Figure~\ref{fig:obs_vtau})
especially when the peak occurs at $v > -400 \mkms$.  One is simply
not sensitive to opacity at $v \sim 0 \mkms$ with the \ion{Mg}{2}
transitions.  This includes gas realted tothe ISM of the galaxy and
also infalling material with modest speeds.  One must take careful
consideration of line emission, therefore,  before drawing robust
conclusions aobut the presence/absence of gas with small velocity
offsets from systemic.

@@

>>>>

Dust is frequently invoked to explain the suppression of line emission
related toresonantly trapped transitions (e.g.\ \lya).  Indeed, a photon
that is trapped for many scatterings within a dusty medium will 
preferentially be extincted relative to a non-resonant photon.  In
$\S$~\ref{sec:dust}, we examined the effects of dust extinction on 
\ion{Mg}{2} and \ion{Fe}{2} emission.  The general result was a modest
suppression of the line flux of the order of (1+\taud)$^{-1}$ instead of
exp(--\taud). Although the \ion{Mg}{2} photons are resonantly
trapped, they require only one to a few scatterings
to escape the wind model and dust has only a
modest effect.  This results from the moderate opacity of the \ion{Mg}{2}
doublet (e.g.\ relative to \lya) and also the velocity law of the
fiducial wind model.
As we have discussed, in scenarios where the \ion{Mg}{2} photons are
more effectively trapped (i.e.\ for many scatterings), dust suppresses
the emission further.
We also emphasize that the \ion{Fe}{2} resonance photons, especially
\feiia, are converted to non-resonant photons and escape the wind
freely.  Therefore, a wind model that trapped \ion{Mg}{2}
photons for many scatterings would not similarly trap 
\feiid\ photons.  The effects of dust are much 
reduced for \ion{Fe}{2}* and, by inference, any other set of
transitions that are coupled to a fine-structure level (e.g.\
\ion{Si}{2} and \ion{Si}{2}*).  In summary,
dust does reduce the line flux relative to the continuum, but we
conclude that it has only a modest effect on the predicted \ion{Mg}{2}
and \ion{Fe}{2}* emission.

Another factor that may reduce the line-emission is a
non-isotropic wind model. One may lower the flux, for example,  by
removing the backside to the wind (e.g.\ the source itself could
shadow a significant fraction of the backside).  To zeroth order, the line emission scales
roughly as $\Omega/4\pi$ where $\Omega$ is the angular extent of the
wind\footnote{There is also a first order dependence on the actual
  orientation of the wind, with maximal emission when the backside is
  fully viewed.}.  Because we require that the wind 
points toward us, it is difficult to reduce $\Omega$ signficantly
below $2 \pi$.  Therefore, we expect this effect to reduce the
emission by at most a factor of order unity.  
Similarly, an anisotropic source may alter the predicted
line-emission.  For example, if the backside of the galaxy were
brighter/fainter then would one predict brighter/fainter emission
relative to the observed continuum.
To reduce the emission, the brightest regions
of the galaxy would need to be oriented toward Earth.  Although this
would not be a
generic occurance, most spectroscopic samples are magnitude-limited
and therefore biased to detecting galaxies when viewed at their
highest fluxes.

Together, dust and non-isotropic wind models may reduce the line-emission
of \ion{Mg}{2} and \ion{Fe}{2}
by a factor of order unity.  This would generally
not be sufficient, however, to reduce the emission below the typical detection
limits.

There is another (more subtle) effect that in principle will
substantially reduce the observed line emission: slit-loss.  As
described in Figures~\ref{fig:fiducial_ifu_mgii} and
\ref{fig:ISM_ifu}, emission from the wind is spatially extended with
a non-zero surface brightness predicted at large radii. This implies 
a non-negligible luminosity emitted well beyond the galaxy.  
The majority of observations of star-forming galaxies to date have
been taken through a spectroscopic slit designed to cover
the brightest continuum regions of the galaxy. Slits with angular
extents of $1-1.5''$ subtend roughly $8-15$\,kpc for galaxies at $z>0.5$.
Therefore, a $1''$ slit covering a galaxy with our fiducial wind would cover
less than half the wind.  The result will be reduced line emission,
even when compared to the galaxy continuum.

We explore the effects of slit-loss as follows.
We model the slit as a perfect, infinitely-long rectangle centered on the
galaxy.  We then tabulate the equivalent width of the line-emission
through slits with a range of widths, parameterized by twice the radius
where the wind has a Sobolev optical depth of $\tau^S = 0.2$ (\rtt).
The line-emission will be weak beyond
this radius because the photons have only a low probability of
interacting with the wind.  For the fiducial wind model, $\mrtt
\approx 15$\,kpc for the \mgiia\ transition (Figure~\ref{fig:fiducial_nvt}).

The predicted equivalent width in emission \ewe\ relative to the absorption
equivalent width \ewabs\ is presented in
Figure~\ref{fig:obs_slit} for a series of transitions for the fiducial
wind model (black curves).
[Add LBG, radiative]
The curves rise very steeply with increasing slit width and then
plateau when the slit width is $\approx \mrtt$.  
The results are comparable for other wind models considered
in this paper (colored curves in Figure~\ref{fig:obs_slit}).
All of the curves rise so steeply that 
the effect is of slit-loss is of order unity
unless the slit-width is very small.   
Nevertheless, the results do motivate
extended aperture observations, e.g.\ integral field unit (IFU)
instrumentation that would map the wind both spatially and spectrally.

Although several effects can reduce the line-emission
relative to the absorption of the outflow, our analysis
indicates that detectable line-emission should occur. 
A principal effect of this emission is
that it modifies the observed absorption profiles (e.g.\
Figure~\ref{fig:noemiss}). This may lead to erroneous conclusions
regarding several characteristics of the flow.  {\it An anaaaalysis of cool gas outflows
that ignores line-emission may incorrectly conclude that the gas is
partially covered, has a significantly lower peak optical depth,
and/or that a $\mdv \sim 0 \mkms$ component is
absent.}  Figure~\ref{fig:obs_ew} demonstrates one observational
consequence of the line emission: reduced measurements for the absorption
equivalent width \ewabs\ of the flow.  In the case of \ion{Mg}{2},
which has the most affected transitions, the \ewabs\ is reduced by
$30-50\%$ from the intrinsic absorption.  In turn, one may derive 
a systematically lower optical depth or velocity extent for the wind,
and therefore a lower total mass and kinetic energy.  
The effects are most pronounced for wind
scenarios where the peak optical depth occurs near $\mdv = 0 \mkms$.
Geometric projection limits the majority of scattered photon emission to
have $|\mdv| < 200 \mkms$; therefore,  the absorption profiles
are filled-in primarily at these velocities.  In
Figure~\ref{fig:obs_peaktau}, we plot the observed peak optical depth
$tau_{\rm pk}$ for \mgiia\ 
versus the velocity where the profile has greatest depth 
for the various wind models (i.e.\
$\tau_{\rm pk}$ vs.\ $v_\tau$ from Tables~\ref{tab:line_diag} and
\ref{tab:plaw_diag}).   In every one of the models, the true peak
optical depth $\tau_{\rm pk}^{\rm true} > 10$.  For the
majority of cases with $v_\tau > -300\mkms$, one observes $\tau_{\rm pk} <
2$ and would infer the wind is not even optically thick!
This occurs because photons scattered by the wind have `filled-in' the
absorption at velocity $\mdv \approx 0 \mkms$.  In contrast, the wind
models with $v_\tau < -400 \mkms$ all give $\tau_{\rm pk} > 3$.  The
results are similar for the \mgiib\ profile.  In fact, one infers a
very similar $\tau_{\rm pk}$ for the two lines which would generally
lead to the false conclusion of partial covering (and, correctly, a
higher $\tau_{\rm pk}^{\rm true}$).  [Last thought?\

All of these (somewhat negative) effects for \ion{Mg}{2} absorption are reduced
for the \ion{Fe}{2} absorption profiles, especially for \feiia.  This
is because a signficant fraction (even a majority) of the
\ion{Fe}{2} emission is florescent \ion{Fe}{2}* emission at longer
wavelengths, which has no effect on the absorption profiles. 
Therefore, the \ion{Fe}{2} absorption equivalent widths (\ewabs) more
closely follow the intrinsic value, one derives more accurate peak
optical depths, and the absorption kinematics more faithfully reflect
the motions of the flow.  Regarding the last point, one also
predicts a velocity offset between the \ion{Fe}{2} and \ion{Mg}{2}
centroids (Figure~\ref{fig:obs_vtau}).
One is simply
less sensitive to absorption at $v \sim 0 \mkms$ with the \ion{Mg}{2}
transitions.  This includes gas realted to the ISM of the galaxy and
also material infalling at a modest speed.  One must take careful
consideration of line emission, therefore,  before drawing robust
conclusions aobut the presence/absence of gas with small velocity
offsets from systemic. [Comment on MgII moving too fast]

@@

[Relative strengths of MgII 2796 and 2803 speak to $\tau$ and
$v_r^{max}$]
